---
title: 'Loan Policy'
---

```{r}
# Load required packages & set global options

library(tidyverse)
library(dplyr)
library(caret)
library(FSelectorRcpp)
library(effectsize)

# Set global options
options(width=100)
```

## Read data

```{r}
# Load in csv data file
loan_data <- read.csv("early_2012_2013_loan_sample_with_outcome.csv")
```

## Data Dictionary

| Variable                    | Description                                                                                                        |
| --------------------------- | ------------------------------------------------------------------------------------------------------------------ |
| id                          | Unique identifier for the loan listing                                                                             |
| member_id                   | Unique identifier for the borrower member                                                                          |
| loan_amnt                   | The listed amount of the loan applied for by the borrower                                                          |
| funded_amnt                 | The total amount committed to the loan at that point in time                                                       |
| funded_amnt_inv             | The total amount committed by investors for that loan at that point in time                                        |
| term                        | The number of payments on the loan (36 or 60 months)                                                               |
| int_rate                    | Interest rate on the loan (%)                                                                                      |
| installment                 | The monthly payment owed by the borrower if the loan originates                                                    |
| grade                       | Lending Club assigned loan grade (A–G)                                                                             |
| sub_grade                   | Lending Club assigned loan subgrade (e.g., A1, A2, B1, etc.)                                                       |
| emp_title                   | The job title supplied by the borrower when applying for the loan                                                  |
| emp_length                  | Employment length in years (0–10+, where 10+ means 10 or more years)                                               |
| home_ownership              | Home ownership status (RENT, OWN, MORTGAGE, OTHER)                                                                 |
| annual_inc                  | The self-reported annual income provided by the borrower                                                           |
| verification_status         | Indicates if income was verified, not verified, or source verified                                                 |
| issue_d                     | The month and year the loan was funded/issued                                                                      |
| loan_status                 | Current status of the loan (e.g., Fully Paid, Charged Off, Current, Default)                                       |
| pymnt_plan                  | Indicates if a payment plan has been put in place for the loan (Y/N)                                               |
| desc                        | Loan description provided by the borrower                                                                          |
| purpose                     | A category provided by the borrower for the loan request (e.g., debt_consolidation, credit_card, home_improvement) |
| title                       | The loan title provided by the borrower                                                                            |
| zip_code                    | The first 3 digits of the zip code provided by the borrower                                                        |
| addr_state                  | The state provided by the borrower in the loan application                                                         |
| dti                         | Debt-to-income ratio: monthly debt payments divided by monthly income                                              |
| delinq_2yrs                 | Number of 30+ days past-due incidences in the past 2 years                                                         |
| earliest_cr_line            | The month and year the borrower's earliest reported credit line was opened                                         |
| inq_last_6mths              | Number of credit inquiries in the past 6 months (excluding auto/mortgage)                                          |
| mths_since_last_delinq      | Number of months since the borrower's last delinquency                                                             |
| mths_since_last_record      | Number of months since the last public record                                                                      |
| open_acc                    | Number of open credit lines in the borrower's credit file                                                          |
| pub_rec                     | Number of derogatory public records (bankruptcies, tax liens, judgments)                                           |
| revol_bal                   | Total credit revolving balance                                                                                     |
| revol_util                  | Revolving line utilization rate (amount of credit used relative to total revolving credit available)               |
| total_acc                   | Total number of credit lines currently in the borrower's credit file                                               |
| initial_list_status         | Initial listing status of the loan (W = Whole, F = Fractional)                                                     |
| out_prncp                   | Remaining outstanding principal for the total amount funded                                                        |
| out_prncp_inv               | Remaining outstanding principal for the portion funded by investors                                                |
| total_pymnt                 | Payments received to date for the total amount funded                                                              |
| total_pymnt_inv             | Payments received to date for the portion funded by investors                                                      |
| total_rec_prncp             | Principal received to date                                                                                         |
| total_rec_int               | Interest received to date                                                                                          |
| total_rec_late_fee          | Late fees received to date                                                                                         |
| recoveries                  | Post charge-off gross recovery                                                                                     |
| collection_recovery_fee     | Post charge-off collection fee                                                                                     |
| last_pymnt_d                | Date of last payment received                                                                                      |
| last_pymnt_amnt             | Amount of last payment received                                                                                    |
| next_pymnt_d                | Next scheduled payment date                                                                                        |
| last_credit_pull_d          | Date Lending Club most recently pulled credit for this loan                                                        |
| collections_12_mths_ex_med  | Number of collections in the past 12 months excluding medical collections                                          |
| mths_since_last_major_derog | Months since most recent 90-day or worse rating                                                                    |
| policy_code                 | Publicly available policy code (1 = new products not publicly available, 2 = public)                               |
| application_type            | Indicates whether the loan is an individual or joint application                                                   |
| acc_now_delinq              | Number of accounts on which the borrower is currently delinquent                                                   |
| tot_coll_amt                | Total collection amounts ever owed                                                                                 |
| tot_cur_bal                 | Total current balance of all accounts                                                                              |
| total_credit_rv             | Total revolving credit limit/high credit                                                                           |
| loan_is_bad                 | Binary target variable indicating if the loan defaulted or was charged off (1 = bad, 0 = good)                     |

## Basic data quality & integrity check

```{r}
# Show head of data
head(loan_data,10)

# Structure of dataset
str(loan_data)

# Summary statistics for numeric and categorical variables
summary(loan_data)

# Count missing values per column
# No major issues identified
colSums(is.na(loan_data))

# Identify duplicate rows -> remove where necessary
# No duplicates identified
duplicates <- loan_data[duplicated(loan_data), ]
nrow(duplicates)

# Drop of current loans
loan_data <- loan_data %>%
  filter(loan_status != "Current")
  
# Create bad_flag variable

bad_status <- c("Charged Off", "Default", "Late (31-120 days)", "Late (16-30 days)", "In Grace Period")

loan_data <- loan_data %>%
  mutate(bad_flag = case_when(
      loan_status %in% bad_status ~ 1,
      loan_status == "Fully Paid" ~ 0,
      TRUE ~ NA_real_ ))

# Interest Rate cleanup (remove %)
if(is.character(loan_data$int_rate)) {
  loan_data$int_rate <- as.numeric(gsub("%", "", loan_data$int_rate))
}

# Convert character predictors to factors
loan_data <- loan_data %>%
  mutate(
    # Handle emp_length NAs as a distinct category
    emp_length = ifelse(is.na(emp_length), "Missing", as.character(emp_length)),
    grade = factor(grade, levels = c("A", "B", "C", "D", "E", "F", "G")),
    sub_grade = as.factor(sub_grade),
    term = as.factor(term),
    emp_length = as.factor(emp_length),
    home_ownership = as.factor(home_ownership))
```

## Check and determine outliers

```{r}
#N/A
```

## Remove outliers and create new data frame

```{r}
#N/A
```

## Create summary data table

```{r}
#N/A
```

## calculation metrics

```{r}
# Net Profit = Total Payment - Funded Amount
loan_data$net_profit <- loan_data$total_pymnt - loan_data$funded_amnt
# ROI = Net Profit / Funded Amount
loan_data$roi <- loan_data$net_profit / loan_data$funded_amnt

# Check ROI range reasonableness
# No extreme values found
summary(loan_data$roi)

# Create the Histogram
ggplot(loan_data, aes(x = roi)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 1) +
  
  # Labels and Titles
  labs(
    title = "Distribution of Loan ROI",
    subtitle = "Dashed line represents breakeven (0% Profit)",
    x = "Return on Investment (ROI)",
    y = "Number of Loans"
  ) +
  
  # Clean Theme for Reports
  theme_minimal() +
  scale_x_continuous(labels = scales::percent)
```

## Calculate information gain

```{r}
# Select only relevant predictors to avoid noise
ig_data <- loan_data %>%
  select(bad_flag, grade, sub_grade, term, loan_amnt, emp_length, home_ownership, annual_inc)

str(ig_data)

# Calculate pure Information Gain
ig_weights <- information_gain(formula = bad_flag ~ grade + sub_grade + term + loan_amnt + emp_length + annual_inc,
                               data = ig_data)
ig_weights

# Format the output
# FSelectorRcpp automatically outputs a clean dataframe with 'attributes' and 'importance'
ig_weights_df <- ig_weights %>%
  arrange(desc(importance))

#Plot 1: Information Gain
plot_ig <- ggplot(ig_weights_df, aes(x = reorder(attributes, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "firebrick") +
  coord_flip() +
  labs(title = "Information Gain: Predicting Loan Defaults",
       subtitle = "Calculated via Shannon Entropy (FSelectorRcpp)",
       x = "Variable", y = "Information Gain (Bits)") +
  theme_minimal()

plot_ig
```

## Policy Application

```{r}

# Add columns to show if each loan would be approved under each policy
loan_data <- loan_data %>%
  mutate(
    approved_A = ifelse(grade %in% c("A", "B"), 1, 0),
    approved_B = ifelse(grade %in% c("A", "B", "C", "D"), 1, 0))

# Filter the full dataset twice based on those new columns to create comparison groups
policy_A_results <- loan_data %>%
  filter(approved_A == 1) %>%
  mutate(policy = "Policy A")

policy_B_results <- loan_data %>%
  filter(approved_B == 1) %>%
  mutate(policy = "Policy B")

# Combine results for comparative analysis
full_comparison <- bind_rows(policy_A_results, policy_B_results)

# Create 4 buckets based on the quartiles of funded_amnt to check OEC robustness
full_comparison$amnt_bucket <- cut(full_comparison$funded_amnt,
                                   breaks = quantile(full_comparison$funded_amnt, probs = seq(0, 1, 0.25)),
                                   include.lowest = TRUE, labels = FALSE)
```

## Summary stat and  t-test

```{r}
# Summary Table
summary_stats <- full_comparison %>%
  group_by(policy) %>%
  summarise(
    Loans_Approved = n(),
    Avg_Loan_Size = mean(funded_amnt),
    Total_Profit = sum(net_profit),
    Avg_ROI = mean(roi),
    Default_Rate = mean(bad_flag))

print("A/B Test Results:")
summary_stats

# Approval rates 
approval_stats <- loan_data %>%
  summarise(
    approval_rate_A = mean(approved_A),
    approval_rate_B = mean(approved_B))

print("Approval Rates:")
approval_stats

# Confidence intervals for ROI
roi_confidence <- full_comparison %>%
  group_by(policy) %>%
  summarise(
    mean_roi = mean(roi),
    ci_lower = t.test(roi)$conf.int[1],
    ci_upper = t.test(roi)$conf.int[2])

print("ROI Confidence Intervals:")
roi_confidence

# Balance check to ensure comparability of policy portfolios
balance_check <- full_comparison %>%
  group_by(policy) %>%
  summarise(
    mean_income = mean(annual_inc, na.rm = TRUE),
    mean_loan_amount = mean(loan_amnt, na.rm = TRUE),
    mean_dti = mean(dti, na.rm = TRUE))

print("Balance Check Between Policies:")
balance_check

# T-Test on ROI (Is the difference in profitability significant?)
t_test_result <- t.test(roi ~ policy, data = full_comparison)

print("Statistical Test (Difference in ROI):")
t_test_result

# Calculate Cohen's d for ROI between Policy B and Policy A
cohens_d(full_comparison$roi ~ full_comparison$policy)

ggplot(full_comparison, aes(x = roi, fill = policy)) +
  geom_histogram(binwidth = 0.05, color = "white", alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ policy) +
  theme_minimal() +
  labs(
    title = "ROI Distribution: Policy A vs Policy B",
    subtitle = "Policy B has more defaults (left spike) but more volume (total area)",
    x = "Return on Investment (ROI)",
    y = "Count of Loans")
```

## Stratification Robustness Check

```{r}
# Group by policy and loan amount bucket to compare Mean ROI (OEC)
robustness_check <- full_comparison %>%
  group_by(policy, amnt_bucket) %>%
  summarise(
    n = n(),
    Mean_ROI = mean(roi),
    .groups = 'drop')

# View the results
print("OEC Robustness Check by Loan Amount Bucket:")
robustness_check
```
